Adaptive group-regularized logistic elastic net regression

### 2019-03-06 docs/manuscript.Rnw
added briers to rnaseq_lymph_node_metastasis.R and micrornaseq_colorectal_cancer.R

  To estimate performance, we split the data into 61 training and 27 test 
  instances, stratified by treatment response. We estimated the models on the 
  training data and calculated AUC on the test data. We present AUC for a range 
  of model sizes, together with the estimated 
  penalty multipliers for \texttt{gren} and \texttt{GRridge} in Figure
  \ref{fig:barplot_lines_micrornaseq_colorectal_cancer_res2_auc}. More 
  performance measures are presented in Section [! section ref here]
  of the SM. In addition, we investigated the sensitivity of
  the multiplier estimation using in SM Section 
  [! section ref here]. We experienced numerical issues with 
  \texttt{GRridge} when cross-validating 
  the global penalty parameter on the training data. We therefore estimated this
  global parameter on the full data for \texttt{GRridge}.

  The penalty multipliers estimated by 
  \texttt{gren} and \texttt{GRridge} are according to expectation: the highly
  expressed group receives the lowest penalty, followed by the medium expressed
  group, while the non-expressed group receives the strongest penalty 
  (Figure \ref{fig:barplot_lines_micrornaseq_colorectal_cancer_res2_auc}a). 
  \texttt{gren} seems to outperform the other methods for a range of model 
  sizes in terms of AUC 
  (Figure \ref{fig:barplot_lines_micrornaseq_colorectal_cancer_res2_auc}b). 
  This is especially pronounced for the larger models. The performance of
  \texttt{cMCP} is somewhat unstable. We noticed that this is in general the 
  case for the group-lasso and its extensions (see SM [! section ref here]). 
  Overall, it seems as though the inclusion of the extra information on the 
  features benefited predictive performance: both \texttt{GRridge} and 
  \texttt{gren} outperform their respective non-group-regularized versions, 
  albeit it only slightly for \texttt{GRridge}.
<<barplot_lines_micrornaseq_colorectal_cancer_res2_auc, fig.cap="Results for the colorectal cancer example with (a) the estimated penalty multipliers and (b) estimated AUC.", out.width="100%", fig.asp=1/2>>=
@

@@

<<>>=
bench1 <- read.table("results/micrornaseq_colorectal_cancer_bench2.csv", 
                     stringsAsFactors=FALSE)
bench2 <- read.table("results/rnaseq_lymph_node_metastasis_bench1.csv", 
                     stringsAsFactors=FALSE)
times1 <- round(bench1$time*10^(-9), 2)
times2 <- round(bench2$time*10^(-9), 2)
session <- gsub("_", ".", devtools::session_info()$platform$system)
@
  We have investigated the computation time of the algorithm and compared to the
  other methods introduced in Sections \ref{sec:colorectal}--\ref{sec:lymphnode}
  on the respective datasets. All timings were obtained on a Macbook Pro 2016 
  running \Sexpr{session}. 
  The timings of \texttt{gren} are competitive. On the colorectal cancer example
  we achieved the following times:
  \Sexpr{times1[c(1:3)]} seconds on the colorectal cancer example, 
  with $\alpha \in \{ 0.05, 0.5, 0.95 \}$, respectively. \texttt{GRridge} is 
  slightly faster with \Sexpr{times1[4]} seconds, while \texttt{SGL} was slightly
  slower: \Sexpr{times1[c(5:7)]} seconds, with 
  $\alpha \in \{ 0.05, 0.5, 0.95 \}$.
  \texttt{cMCP} (\Sexpr{times1[c(8:10)]} seconds) and \texttt{gel} 
  (\Sexpr{times1[c(11:13)]} seconds) are faster than \texttt{gren}. 
  On the lymph node cancer example \texttt{gren} achieved 
  \Sexpr{times2[c(1:3)]}, with $\alpha \in \{ 0.05, 0.5, 0.95 \}$, respectively.
  \texttt{GRridge} was slightly faster with \Sexpr{times2[4]} seconds, 
  while \texttt{SGL} was slightly slower: \Sexpr{times2[c(5:7)]} seconds, with 
  $\alpha \in \{ 0.05, 0.5, 0.95 \}$.
  \texttt{cMCP} (\Sexpr{times2[c(8:10)]} seconds) and \texttt{gel} 
  (\Sexpr{times2[c(11:13)]} seconds) are faster than \texttt{gren}. The 
  overlapping group methods achieved \Sexpr{times2[c(14:16)]} seconds
  (\texttt{ocMCP}) and \Sexpr{times2[c(17:19)]} seconds (\texttt{ogel}), for 
  $\alpha \in \{ 0.05, 0.5, 0.95 \}$, respectively.
  We note that a considerable speedup may be achieved by 
  doing one EB update for every few VB updates, instead of running the VB EM 
  until convergence. This works in practice, because an EM tends to take a few 
  initial large steps towards the optimum, followed by a series of smaller 
  steps. We argue that The first few iterations of the VB EM render the 
  approximate expectation in [! equation ref here] accurate enough for a 
  reliable EB step. In addition, this is the approach often taken by more
  computationally oriented disciplines. For example, in many machine learning 
  domains, the problem is viewed as purely algorithmic [! references here]. The
  the EB and VB 
  steps optimisation steps are interpreted as steps along the gradient in the 
  larger combined space of model- and hyper-parameters.

>>>>

  To estimate performance, we split the data into 61 training and 27 test 
  instances, stratified by treatment response. We estimated the models on the 
  training data and calculated AUC on the test data. We present AUC for a range 
  of model sizes, together with the estimated 
  penalty multipliers for \texttt{gren} and \texttt{GRridge} in Figure
  \ref{fig:barplot_lines_micrornaseq_colorectal_cancer_res1_auc}. More 
  performance measures are presented in Section [! section ref here]
  of the SM. In addition, we investigated the sensitivity of
  the multiplier estimation using in SM Section 
  [! section ref here]. We experienced numerical issues with 
  \texttt{GRridge} when cross-validating 
  the global penalty parameter on the training data. We therefore estimated this
  global parameter on the full data for \texttt{GRridge}.

  The penalty multipliers estimated by 
  \texttt{gren} and \texttt{GRridge} are according to expectation: the highly
  expressed group receives the lowest penalty, followed by the medium expressed
  group, while the non-expressed group receives the strongest penalty 
  (Figure \ref{fig:barplot_lines_micrornaseq_colorectal_cancer_res1_auc}a). 
  \texttt{gren} seems to outperform the other methods for a range of model 
  sizes in terms of AUC 
  (Figure \ref{fig:barplot_lines_micrornaseq_colorectal_cancer_res1_auc}b). 
  This is especially pronounced for the larger models. The performance of
  \texttt{cMCP} is somewhat unstable. We noticed that this is in general the 
  case for the group-lasso and its extensions (see SM [! section ref here]). 
  Overall, it seems as though the inclusion of the extra information on the 
  features benefited predictive performance: both \texttt{GRridge} and 
  \texttt{gren} outperform their respective non-group-regularized versions, 
  albeit it only slightly for \texttt{GRridge}.
<<barplot_lines_micrornaseq_colorectal_cancer_res1_auc, fig.cap="Results for the colorectal cancer example with (a) the estimated penalty multipliers and (b) estimated AUC.", out.width="100%", fig.asp=1/2>>=
@

@@

<<>>=
bench1 <- read.table("results/micrornaseq_colorectal_cancer_bench1.csv", 
                     stringsAsFactors=FALSE)
bench2 <- read.table("results/rnaseq_lymph_node_metastasis_bench1.csv", 
                     stringsAsFactors=FALSE)
times1 <- round(bench1$time*10^(-9), 2)
times2 <- round(bench2$time*10^(-9), 2)
session <- gsub("_", ".", devtools::session_info()$platform$system)
@
  We have investigated the computation time of the algorithm and compared to the
  other methods introduced in Sections \ref{sec:colorectal}--\ref{sec:lymphnode}
  on the respective datasets. All timings were obtained on a Macbook Pro 2016 
  running \Sexpr{session}. 
  The timings of \texttt{gren} are competitive. On the colorectal cancer example
  we achieved the following times:
  \Sexpr{times1[c(1:3)]} seconds on the colorectal cancer example, 
  with $\alpha \in \{ 0.05, 0.5, 0.95 \}$, respectively. \texttt{GRridge} is 
  slightly faster with \Sexpr{times1[4]} seconds, while \texttt{SGL} was slightly
  slower: \Sexpr{times1[c(5:7)]} seconds, with 
  $\alpha \in \{ 0.05, 0.5, 0.95 \}$.
  \texttt{cMCP} (\Sexpr{times1[c(8:10)]} seconds) and \texttt{gel} 
  (\Sexpr{times1[c(11:13)]} seconds) are faster than \texttt{gren}. 
  On the lymph node cancer example \texttt{gren} achieved 
  \Sexpr{times2[c(1:3)]}, with $\alpha \in \{ 0.05, 0.5, 0.95 \}$, respectively.
  \texttt{GRridge} was slightly faster with \Sexpr{times2[4]} seconds, 
  while \texttt{SGL} was slightly slower: \Sexpr{times2[c(5:7)]} seconds, with 
  $\alpha \in \{ 0.05, 0.5, 0.95 \}$.
  \texttt{cMCP} (\Sexpr{times2[c(8:10)]} seconds) and \texttt{gel} 
  (\Sexpr{times2[c(11:13)]} seconds) are faster than \texttt{gren}. The 
  overlapping group methods achieved \Sexpr{times2[c(14:16)]} seconds
  (\texttt{ocMCP}) and \Sexpr{times2[c(17:19)]} seconds (\texttt{ogel}), for 
  $\alpha \in \{ 0.05, 0.5, 0.95 \}$, respectively.
  We note that a considerable speedup may be achieved by 
  doing one EB update for every few VB updates, instead of running the VB EM 
  until convergence. This works in practice, because an EM tends to take a few 
  initial large steps towards the optimum, followed by a series of smaller 
  steps. We argue that The first few iterations of the VB EM render the 
  approximate expectation in [! equation ref here] accurate enough for a 
  reliable EB step. In addition, this is the approach often taken by more
  computationally oriented disciplines. For example, in many machine learning 
  domains, the problem is viewed as purely algorithmic [! references here]. The
  the EB and VB 
  steps optimisation steps are interpreted as steps along the gradient in the 
  larger combined space of model- and hyper-parameters.
